Assessment of zonally symmetric and asymmetric components of the Southern Annular Mode using a novel approach

### 2021-05-28 analysis/paper/asymsam.tex
Render

The method assumes linearity in the asymmetric component of the SAM.
That means that zonal asymmetries associated with positive SAM phase (SAM+) are almost opposite in sign and of the same magnitude to the ones associated with negative SAM phase (SAM-).
\citet{fogt2012}'s composites (their Figure 4) suggest that this might not be entirely valid, although much of that apparent non-linearity could be due to the heterogenous nature of the selected years for constructing the composites.
To test this assumption, we computed seasonal composites of zonal anomalies of geopotential height for SAM+ and SAM- (defined as months in which the SAM index is greater than 1 standard deviation and lower than minus 1 standard deviation, respectively) for the period from 1979 to 2018 at the 700~hPa and 50~hPa levels (Figures \ref{fig:A3} and \ref{fig:A4}).
In all seasons and both levels, SAM+ composites are similar to SAM- in structure but with the opposite sign.
Spatial correlations between composites for each season are high.
The method considered in this study seems then a reasonable approximation of the phenomenon.

@@

The wave-3 pattern from Figure~\ref{fig:2d-regr}b.2 is very similar to the PSA Pattern \citep{mo1987, kidson1988} which is a teleconnection pattern associated with the ENSO \citep{karoly1989}.
Indeed, \citet{fogt2011} showed that there is a significant relationship between the SAM and the ENSO.
The correlation between the SAM and the ENSO as measured by the Oceanic Niño Index \citep[ONI,][]{bamston1997} is shown in Table~\ref{tab:enso-cor-table} for each SAM index and for each trimester.
There is a significant correlation between SAM and ENSO. When divided in trimesters, this correlation is important only in DJF and SON.
This relationship is captured mainly by the A\nobreakdash-SAM, as this index has signifcant partial correlations with ENSO while correlations with S\nobreakdash-SAM are all smaller and non-significant.
In MAM, ENSO is not significantly correlated with SAM, but it is significantly correlated with A\nobreakdash-SAM in a level comparable with the ENSO-SAM correlation in SON. The same analysis was performed using the Multivariate ENSO Index \citep{wolter2011} and the Southern Oscillation Index \citep{ropelewski1987}, obtaining similar results.
The latter allows us to conclude that these results do not depend on the ENSO index used.

@@

In Autumn, Winter and Spring (rows 2, 3, and 4 in Figure \ref{fig:regr-air-season}) the positive ring is only present through its local maximums in the regression with the SAM, which reflects the more asymmetric nature of the SAM compared to Summer.
There are also negative anomalies in Southern Australia, and positive anomalies over New Zealand and Southern South America.
Similar features were observed in station measurements by \citet{jones2019}, although using data from 1957 to 2016.
In Spring, the tropical signal of A\nobreakdash-SAM is similar to the one in Summer, again revealing the significance of the ENSO-A\nobreakdash-SAM link.
Moreover, from Autumn to Spring, positive (negative) temperature anomalies are discernible at the northern (southern) portion of the Antarctic Peninsula in the regressions with the SAM, which are also evident in the regressions with A-SAM. In agreement, \citet{marshall2016} found the same signal in association with SAM while that pattern is only evident in Autumm for the southern baroclinic annular mode, which is associated with variations in extratropical storm amplitude.

@@

Figure~\ref{fig:global-pp} shows regression of the SAM indices with precipitation for the Southern Hemisphere. The precipitation signal associated with SAM shows generally decreased precipitation at around 45ºS, somewhat increase precipitation around 30ªS (Fig.~\ref{fig:global-pp}a) and increased precipitation over Antarctica, a pattern known from other studies \citep[e.g.][]{hendon2014}.
This pattern is mostly unchanged between seasons albeit varying in intensity (not shown).
Panels b and c in Figure~\ref{fig:global-pp} shows that the A\nobreakdash-SAM signal is only in the tropics and mid-latitudes, while the S\nobreakdash-SAM signal is strong in the higher latitudes.
In particular positive values of S\nobreakdash-SAM is associated with the increased precipitation over Antarctica and decrease precipitation around the Southern Ocean.

>>>>

The method assumes linearity in the asymmetric component of the SAM.
That means that zonal asymmetries associated with positive SAM phase (SAM+) are almost opposite in sign and of the same magnitude to the ones associated with negative SAM phase (SAM-).
\citet{fogt2012}'s composites (their Figure 4) suggest that this might not be entirely valid, although much of that apparent non-linearity could be due to the heterogeneous nature of the selected years for constructing the composites.
To test this assumption, we computed seasonal composites of zonal anomalies of geopotential height for SAM+ and SAM- (defined as months in which the SAM index is greater than 1 standard deviation and lower than minus 1 standard deviation, respectively) for the period from 1979 to 2018 at the 700~hPa and 50~hPa levels (Figures \ref{fig:A3} and \ref{fig:A4}).
In all seasons and both levels, SAM+ composites are similar to SAM- in structure but with the opposite sign.
Spatial correlations between composites for each season are high.
The method considered in this study seems then a reasonable approximation of the phenomenon.

@@

The wave-3 pattern from Figure~\ref{fig:2d-regr}b.2 is very similar to the PSA Pattern \citep{mo1987, kidson1988} which is a teleconnection pattern associated with the ENSO \citep{karoly1989}.
Indeed, \citet{fogt2011} showed that there is a significant relationship between the SAM and the ENSO.
The correlation between the SAM and the ENSO as measured by the Oceanic Niño Index \citep[ONI,][]{bamston1997} is shown in Table~\ref{tab:enso-cor-table} for each SAM index and for each trimester.
There is a significant correlation between SAM and ENSO. When divided in trimesters, this correlation is important only in DJF and SON.
This relationship is captured mainly by the A\nobreakdash-SAM, as this index has significant partial correlations with ENSO while correlations with S\nobreakdash-SAM are all smaller and non-significant.
In MAM, ENSO is not significantly correlated with SAM, but it is significantly correlated with A\nobreakdash-SAM in a level comparable with the ENSO-SAM correlation in SON. The same analysis was performed using the Multivariate ENSO Index \citep{wolter2011} and the Southern Oscillation Index \citep{ropelewski1987}, obtaining similar results.
The latter allows us to conclude that these results do not depend on the ENSO index used.

@@

In Autumn, Winter and Spring (rows 2, 3, and 4 in Figure \ref{fig:regr-air-season}) the positive ring is only present through its local maximums in the regression with the SAM, which reflects the more asymmetric nature of the SAM compared to Summer.
There are also negative anomalies in Southern Australia, and positive anomalies over New Zealand and Southern South America.
Similar features were observed in station measurements by \citet{jones2019}, although using data from 1957 to 2016.
In Spring, the tropical signal of A\nobreakdash-SAM is similar to the one in Summer, again revealing the significance of the ENSO-A\nobreakdash-SAM link.
Moreover, from Autumn to Spring, positive (negative) temperature anomalies are discernible at the northern (southern) portion of the Antarctic Peninsula in the regressions with the SAM, which are also evident in the regressions with A-SAM. In agreement, \citet{marshall2016} found the same signal in association with SAM while that pattern is only evident in Autumn for the southern baroclinic annular mode, which is associated with variations in extratropical storm amplitude.

@@

Figure~\ref{fig:global-pp} shows regression of the SAM indices with precipitation for the Southern Hemisphere. The precipitation signal associated with SAM shows generally decreased precipitation at around 45\degree~S, somewhat increase precipitation around 30\degree\S (Fig.~\ref{fig:global-pp}a) and increased precipitation over Antarctica, a pattern known from other studies \citep[e.g.][]{hendon2014}.
This pattern is mostly unchanged between seasons albeit varying in intensity (not shown).
Panels b and c in Figure~\ref{fig:global-pp} shows that the A\nobreakdash-SAM signal is only in the tropics and mid-latitudes, while the S\nobreakdash-SAM signal is strong in the higher latitudes.
In particular positive values of S\nobreakdash-SAM is associated with the increased precipitation over Antarctica and decrease precipitation around the Southern Ocean.
